Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-02-19 paper/main.tex
revisions to section 6. added a citation to bib

Note that in the iterative procedure outlined in \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrange function compared to solving the original problem
\eqref{eq:l1min} directly. In fact, the first step that minimizes $L_{\mu_k}
(\xx,\yy_k)$ assuming $\yy_k$ and $\mu_k$ are fixed can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \ref{alg:alm} summarizes the ALM $\ell_1$-min algorithm.

@@

Below is the psuedocode solving the minimization problem posed in \eqref{eq:l1min_denoise}.

\begin{algorithm}[t]
\caption{\bf (Augmented Lagrange Multiplier Method for
Recognition)}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \bb$, $\y_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}\left(\bb - A\x_k +
\frac{1}{\mu}\y_k, \frac{1}{\mu}\right)$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}\left(\z_l +
\frac{1}{\gamma}A^T\left(\bb - A\z_l - \e_{k+1} +
\frac{1}{\mu}\y_k\right), \frac{1}{\mu\gamma}\right)$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}\left( 1 +
\sqrt{1+4t_l^2}\right)$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$,  \; $\y_{k+1} \leftarrow \y_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{small}
\end{algorithmic}
\label{alg:alm_rec} 
\end{algorithm}

@@

\subsection{GPU Implementation}
{\bf AY: A diagram should be used to illustrate how the jobs are distributed on a GPU architecture.}
The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. Because data transfer from the 
RAM to GPU suffers bandwidth limitations, we designed our implementation to transfer the A matrix and measurement vector at once.  
The algorithm is similar to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU keeps all of the data and runs the computation.

@@

In this section, we present four experiments to benchmark the performance of our algorithm on CPU and GPU platforms. 
To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).  Unless otherwise specified, all implementations utilize single
precision floating point datatypes.

The first experiment is designed to compare the two implementations of PALM
that solve a single large instance of the $\ell_1$-minimization problem (Algorithm \ref{alg:alm_rec}).  
The measurement $d \times n$ matrix $A$ was generated as a random Gaussian matrix, with each entry
randomly generated from the standard normal and normalized with unit norm.  The ratio of $d:n$ varied from
$10:1$ to $10:5$ with the dimension of $d$ of matrix $A$ varying from 500 to 8000.
The ground truth signals, $x_0$ and $y_k$ had a sparsity rate of 0.1 with elements generated using the standard normal and normalized with unit norm.
The measurement vector was generated with the equation $\y_0 = A \x_0 + \y_k$.    We set the number of inner iterations at 50 and outer iterations at 5000.
We also varied the different inner and tolerance levels from 1e-3 to 1e-6.
We averaged the the time taken for the ALM algorithm to solve 10 instances of a problem.

The x set of experiments will compare the CPU and GPU versions of the entire pipeline, 
which includes the alignment and recognition stage.

>>>>

Note that in the iterative procedure outlined in \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrange function compared to solving the original problem
\eqref{eq:l1min} directly. In fact, the first step that minimizes $L_{\mu_k}
(\xx,\yy_k)$ assuming $\yy_k$ and $\mu_k$ are fixed can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \ref{alg:alm} summarizes the ALM $\ell_1$-min algorithm \footnote{While we
present the ALM algorithm in the primal domain, there also
exists an implementation ALM in the dual domain.}.

@@

Below is the algorithm for the face recognition stage in Figure \ref{fig:pipeline}, which
solves the $\ell_1$ minimization problem posed in \eqref{eq:l1min_denoise}.

\begin{algorithm}[t]
\caption{\bf (Face Recognition via ALM)}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \bb$, $\y_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}\left(\bb - A\x_k +
\frac{1}{\mu}\y_k, \frac{1}{\mu}\right)$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}\left(\z_l +
\frac{1}{\gamma}A^T\left(\bb - A\z_l - \e_{k+1} +
\frac{1}{\mu}\y_k\right), \frac{1}{\mu\gamma}\right)$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}\left( 1 +
\sqrt{1+4t_l^2}\right)$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$,  \; $\y_{k+1} \leftarrow \y_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{small}
\end{algorithmic}
\label{alg:alm_rec} 
\end{algorithm}

@@

\subsection{GPU Implementation}
{\bf AY: A diagram should be used to illustrate how the jobs are distributed on a GPU architecture.}
The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. Because data transfer from the 
RAM to GPU suffers bandwidth limitations, we designed our implementation to transfer the A matrix and measurement vector once.  
The algorithm is similar to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU keeps all of the data and runs the computation.

@@

In this section, we present four experiments to benchmark the performance of our algorithm on CPU and GPU platforms. 
To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).  Unless otherwise specified, all implementations utilize single
precision floating point datatypes.  For all of the experiments, we set the number 
of inner loop iterations at 50 and outer loop iterations at 5000.

The first experiment is designed to compare the two implementations of PALM
that solve a single large instance of the $\ell_1$-minimization problem (Algorithm \ref{alg:alm_rec}) by
comparing the amount of time to solve a problem given a generic random matrix $A$ and a random sparse
vector $\xx_0$.  
The measurement $d \times n$ matrix $A$ was generated as a random Gaussian matrix, with each entry
randomly generated from the standard normal and normalized with unit norm.  The ratio of $d:n$ varied from
$10:1$ to $10:5$ with the dimension of $d$ of matrix $A$ varying from 500 to 8000.
The ground truth signals, $\xx_0$ and $\yy_k$ had a sparsity rate of 0.1 with elements generated using the standard normal and normalized with unit norm.
Because the ground truth signal $\xx_0$ and $yy_k$ are known, the algorithm terminates
when $norm(\bx-\bx_0) < tolerance$.
The measurement vector was generated with the equation $\y_0 = A \x_0 + \y_k$.   
We also varied the different inner and tolerance levels from 1e-3 to 1e-6.
We averaged the the time taken for the ALM algorithm to solve 10 instances of a problem.

%The x set of experiments will compare the CPU and GPU versions of the entire pipeline, 
%which includes the alignment and recognition stage.

The second experiment is to compare the amount of time to solve the face alignment
stage for each implementation.  Because the alignment stage is not affected by the resolution of the image,
we fix the resolution at 60x80.
{\bf VS:  this accuracy number depends on the number of subjects in the A matrix right?  So we either 
need to fix the number of subjects to compare against or vary it...}

The third experiment is to compare accuracy rate of the face recognition algorithm at different
image resolutions.

The fourth experiment is to compare the speed of the recognition stage w.r.t the number of 
subjects in the matrix $A$.

Our final experiment is to compare the speeds of CPU and GPU implementations entire pipeline in Figure \ref{fig:pipeline}.
